---
title: "The Value Proposition of Stablecoins in South Africa's Evolving Digital Payment Landscape: A Comparative Analysis"
subtitle: "ECO5016W - Minor Dissertation in FinTech (Non-unitary option: Topic 1 of 2)"
author: "Cesaire Tobias"
date: "2026-06-23"
format:
  pdf:
    date-format: "DD MMMM, YYYY"
    toc: true
    toc-depth: 3
    lot: true
    lof: true
    number-sections: true
    colorlinks: true
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
    linestretch: 1.5
    cite-method: citeproc
    bibliography: references.bib
    csl: harvard-cite-them-right.csl
    link-citations: true
    keep-tex: false
    fig-pos: 'H'
    tbl-pos: 'H'
    include-in-header:
      text: |
        \usepackage{float}
        \usepackage{placeins}
        \usepackage{array}
        \usepackage{booktabs}
        \floatplacement{figure}{H}
        \floatplacement{table}{H}
execute:
  echo: false
  warning: false
  message: false
---

```{r setup, include=FALSE}

options(repos = c(CRAN = "https://cran.rstudio.com/"))

required_packages <- c("countrycode", "dplyr", "ggplot2", "readxl", "stringr", "tidyr", "forcats", "flextable", "officer")
optional_packages <- c("cqThemes")

github_packages <- list(
  cqThemes = "ces0491/cqThemes"
)

if (!"remotes" %in% installed.packages()[, "Package"]) {
  install.packages("remotes", quiet = TRUE)
}

installed <- installed.packages()[, "Package"]
missing_packages <- setdiff(required_packages, installed)

cran_missing <- setdiff(missing_packages, names(github_packages))
if (length(cran_missing) > 0) {
  install.packages(cran_missing, quiet = TRUE)
}

for (pkg in optional_packages) {
  if (!pkg %in% installed && pkg %in% names(github_packages)) {
    tryCatch({
      remotes::install_github(github_packages[[pkg]], quiet = TRUE)
    }, error = function(e) NULL)
  }
}

invisible(lapply(required_packages, function(pkg) {
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}))

for (pkg in optional_packages) {
  tryCatch({
    suppressPackageStartupMessages(library(pkg, character.only = TRUE))
  }, error = function(e) NULL)
}

cqthemes_available <- "cqThemes" %in% (.packages())

if (cqthemes_available) {
  tryCatch({
    use_pdf_theme(seed = 0491)
  }, error = function(e) {
    cqthemes_available <<- FALSE
  })
}

if (!cqthemes_available) {
  theme_set(theme_minimal())
}

if (cqthemes_available) {
  get_colors <- function(n = 4, type = "categorical") {
    cq_colors(n, type = type)
  }
  get_theme <- function() theme_cq_pdf()
} else {
  get_colors <- function(n = 4, type = "categorical") {
    colors <- c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", "#8c564b")
    return(colors[1:min(n, length(colors))])
  }
  get_theme <- function() theme_minimal()
}

options(scipen = 999)

```

```{r wordcount, include=FALSE}

count_words_qmd <- function(file_path) {
  raw_text <- readLines(file_path, warn = FALSE)
  
  body_text <- ""
  in_code_chunk <- FALSE
  in_yaml <- FALSE
  in_table <- FALSE
  yaml_delim_count <- 0
  
  for (i in seq_along(raw_text)) {
    line <- raw_text[i]
    line_trimmed <- trimws(line)
    
    # Handle YAML header
    if (line_trimmed == "---") {
      yaml_delim_count <- yaml_delim_count + 1
      if (yaml_delim_count == 1) {
        in_yaml <- TRUE
        next
      } else if (yaml_delim_count == 2) {
        in_yaml <- FALSE
        next
      }
    }
    
    if (in_yaml) next
    
    # Stop at References section
    if (grepl("^#+\\s*(References|Bibliography|Works Cited|Literature Cited)", 
              line_trimmed, ignore.case = TRUE)) {
      break
    }
    
    # Handle code chunks
    if (grepl("^```", line_trimmed)) {
      in_code_chunk <- !in_code_chunk
      next
    }
    
    if (in_code_chunk) next
    
    # Skip headers
    if (grepl("^#+", line_trimmed)) next
    
    # Skip figure/table captions and labels
    if (grepl("^fig\\.|^tbl\\.|fig\\.cap|tbl\\.cap|caption\\s*=|include\\s*=|echo\\s*=", line_trimmed)) next
    
    # Skip table content (looking for flextable patterns)
    if (grepl("flextable\\(|^\\|", line_trimmed)) {
      in_table <- TRUE
      next
    }
    
    # End of table detection (empty line after table)
    if (in_table && line_trimmed == "") {
      in_table <- FALSE
      next
    }
    
    if (in_table) next
    
    processed_line <- line
    
    # Remove inline code
    processed_line <- gsub("`[^`]*`", "", processed_line)
    
    # Remove citations - both @ style and parenthetical
    processed_line <- gsub("\\[@[^]]+\\]", "", processed_line)  # [@Author2021]
    processed_line <- gsub("\\([A-Z][^)]*\\d{4}[a-z]?[^)]*\\)", "", processed_line)  # (Author, 2021)
    processed_line <- gsub("[@][A-Za-z0-9_:-]+", "", processed_line)  # @citations
    
    # Remove markdown formatting but keep text
    processed_line <- gsub("\\*\\*([^*]+)\\*\\*", "\\1", processed_line)  # bold
    processed_line <- gsub("\\*([^*]+)\\*", "\\1", processed_line)  # italics
    processed_line <- gsub("_([^_]+)_", "\\1", processed_line)  # underscore emphasis
    processed_line <- gsub("\\[([^]]+)\\]\\([^)]+\\)", "\\1", processed_line)  # links
    processed_line <- gsub("!\\[([^]]*)\\]\\([^)]+\\)", "", processed_line)  # images
    processed_line <- gsub("<[^>]+>", "", processed_line)  # HTML tags
    
    # Remove footnote markers
    processed_line <- gsub("\\^\\[[^]]+\\]", "", processed_line)
    
    body_text <- paste(body_text, processed_line)
  }
  
  # Clean up punctuation and count words
  body_text <- gsub("[[:punct:]]+", " ", body_text)
  words <- unlist(strsplit(body_text, "\\s+"))
  words <- words[nchar(trimws(words)) > 0]
  
  return(length(words))
}

# Academic word count (excluding tables, citations, etc.)
word_count <- count_words_qmd(knitr::current_input())

# For comparison, create a "total content" count that includes everything
count_all_words <- function(file_path) {
  raw_text <- readLines(file_path, warn = FALSE)
  body_text <- ""
  in_yaml <- FALSE
  in_code <- FALSE
  yaml_delim_count <- 0
  
  for (line in raw_text) {
    if (trimws(line) == "---") {
      yaml_delim_count <- yaml_delim_count + 1
      if (yaml_delim_count == 1) {
        in_yaml <- TRUE
        next
      } else if (yaml_delim_count == 2) {
        in_yaml <- FALSE
        next
      }
    }
    if (in_yaml) next
    if (grepl("^#+\\s*References", line)) break
    if (grepl("^```", trimws(line))) {
      in_code <- !in_code
      next
    }
    if (!in_code) {
      body_text <- paste(body_text, line)
    }
  }
  
  body_text <- gsub("[[:punct:]]+", " ", body_text)
  words <- unlist(strsplit(body_text, "\\s+"))
  words <- words[nchar(trimws(words)) > 0]
  return(length(words))
}

total_word_count <- count_all_words(knitr::current_input())

```

\newpage

# Abstract

This dissertation examines the value proposition of stablecoins within South Africa's evolving payment ecosystem through comparative analysis of digital payment alternatives. The research investigates whether stablecoins offer meaningful advantages over existing regulated digital payment infrastructure or merely replicate established services with additional complexity and risk exposure. Drawing from Brazil's implementation of Pix as a benchmark for effective digital payment system design and utilizing theoretical frameworks from @adrian2021rise and @catalini2022some, this study analyzes stablecoins across technical, economic, and institutional dimensions. The analysis reveals that while stablecoins demonstrate certain advantages in cross-border transactions and programmable payment functionality, their benefits within domestic payment contexts are constrained by the capabilities of modern, well-designed payment infrastructure. The findings suggest that in jurisdictions with robust, interoperable, and inclusive digital payment systems under appropriate regulatory oversight, stablecoins primarily serve specialized use cases rather than providing broad-based improvements to payment efficiency or financial inclusion. This conclusion has significant implications for policymakers, financial institutions, and technology developers operating within emerging economies implementing payment system modernization programmes.

::: {=latex}
\vspace{1\baselineskip}
:::

**Keywords:** stablecoins, digital payments, financial inclusion, payment systems, South Africa, financial technology, regulatory frameworks

::: {=latex}
\vspace{1\baselineskip}
:::

**Word Count: `r formatC(word_count, format="d", big.mark=",")`**

------------------------------------------------------------------------

\newpage

# Introduction

## Research Context and Motivation {#sec-research-context}

Stablecoins have emerged as a prominent category of crypto assets designed to maintain stable value relative to reference currencies or baskets of assets [@adrian2021rise]. Global stablecoin market capitalization exceeded \$190 billion by late 2024, with Tether (USDT) and USD Coin (USDC) accounting for over 90% of market share as of August 2025 [@coinmarketcap2025total]. Unlike traditional cryptocurrencies, stablecoins employ various mechanisms to maintain price stability, including full reserve backing with fiat currencies, over-collateralization with crypto assets, or algorithmic supply adjustments—though the latter approach notably failed with TerraUSD's \$50 billion collapse in May 2022 [@liu2023anatomy]. The emergence of stablecoins reflects broader trends in the digitization of money and payments, driven by technological advances in distributed ledger technology, mobile computing, and cryptographic protocols [@auer2020technology]. Proponents argue that these innovations represent a fundamental advancement in digital payment infrastructure, potentially addressing longstanding inefficiencies in cross-border payments, financial inclusion, and programmable money applications [@catalini2022some; @arner2020stablecoins]. However, whether stablecoins actually address these inefficiencies – particularly when compared to modernized traditional payment systems – remains an empirical question requiring systematic evaluation.

Sophisticated government-backed digital payment platforms in emerging economies present compelling counter-narratives to the necessity of blockchain-based payment solutions. Brazil's Pix system, launched in November 2020 by the Central Bank of Brazil (BCB), exemplifies this trend. Key Pix metrics include: 150 million users within three years, 42 billion transactions in 2023, daily peaks exceeding 200 million transactions, and average costs below \$0.10 per transaction [@bcb2024pix; @duarte2022central]. These benchmarks will serve as comparative reference points throughout this analysis.

The Brazilian experience demonstrates that many benefits commonly attributed to blockchain-based payment systems can be achieved through careful design and implementation of traditional payment infrastructure. While in South Africa[^1], the ongoing payment system modernization, anchored by the South African Reserve Bank's Vision 2025 strategy [@sarb2018vision] and the implementation of PayShap in 2023 [@bankservafrica2024businessreport], raises questions about the residual value proposition of stablecoins in jurisdictions with well-functioning digital payment infrastructure.

[^1]: South Africa is classified as an upper-middle-income economy [@worldbank2024country], with sophisticated financial markets including Africa's largest stock exchange by market capitalization [@sse_initiative_jse_profile], but significant socioeconomic challenges with a Gini coefficient of 0.63, among the highest globally [@statssa2019livingconditions]

## Research Objectives and Questions

This research addresses the comparative advantages and limitations of stablecoins within evolving national payment systems. The primary research question is:

*In countries with high-functioning, low-cost interoperable digital payment systems under public oversight, do stablecoins offer residual value, or do they replicate existing services with added complexity and risk?*

Understanding the relative benefits and constraints of different technological approaches has far-reaching policy implications, informing resource allocation decisions and regulatory frameworks.

## Scope and Approach

This analysis focuses on South Africa as the primary case study, using Brazil's Pix system as a comparative benchmark for evaluating the potential of well-designed traditional payment infrastructure. The research employs comparative analysis incorporating frameworks from monetary economics, payment system design, and financial inclusion research. The research examines stablecoins broadly, encompassing fiat-collateralized (such as USDT and USDC), crypto-collateralized (such as DAI), and algorithmic variants, while recognizing that different stablecoin designs may have different performance characteristics and use cases [@clements2021built].

# Literature Review

## Theoretical Foundations of Digital Payment Systems

The evolution from traditional to digital payment systems has generated extensive theoretical debate about the fundamental nature of money and value transfer mechanisms. A substantial body of literature examines how digitization transforms the core functions of payment systems, with scholars converging on several critical dimensions that determine system effectiveness. Network effects and market structure emerge as dominant themes across the payment systems literature. The two-sided market framework, developed by @rochet2003platform, @rochet2006progress and @armstrong2006competition, demonstrates how payment platforms must simultaneously attract both consumers and merchants, creating complex pricing dynamics and adoption challenges. This theoretical foundation has been assessed across multiple contexts, with consistent findings that achieving critical mass thresholds is essential before self-sustaining growth occurs [@bcg2024global; @chakravorti2006platform]. The implications for stablecoin adoption are particularly relevant, as blockchain-based systems must overcome the same coordination problems without the institutional mechanisms available to traditional systems [@catalini2020some; @cong2021tokenomics].

The settlement finality debate reveals fundamental architectural differences between payment system designs. Traditional payment theory, grounded in the work of @kahn2009why and @norman2010liquidity, emphasizes the importance of immediate and irrevocable settlement through central bank money. This contrasts sharply with the probabilistic finality inherent in blockchain systems, analyzed by @gazi2020tight and @bonneau2015sok. The tension between immediate finality and decentralized consensus represents a trade-off that multiple authors identify as potentially insurmountable for certain use cases [@auer2020technology; @abadi2018blockchain]. The theoretical model developed by @chiu2019blockchain explores these trade-offs, concluding that blockchain-based settlement can impose higher liquidity costs on participants to ensure transactional integrity compared to centralized account-based systems.

Transaction cost economics provides another lens through which scholars analyze payment system efficiency. Building on @williamson1985economic's framework, recent literature examines how different payment architectures affect total transaction costs, including both explicit fees and implicit costs such as time, cognitive burden, and risk [@bolt2011pricing; @arango2015consumer]. The stablecoin literature contributes unique insights here, with @catalini2022some and @lyons2023what demonstrating how blockchain-based systems introduce new cost categories—gas fees[^2], slippage, smart contract[^3] risk—that traditional frameworks inadequately capture. This is particularly evident in transaction fees, which are subject to extreme volatility due to on-chain auction mechanisms that respond to network congestion [@abadi2018blockchain].

[^2]: Transaction fees paid to blockchain network validators for processing and confirming transactions, varying based on network congestion

[^3]: Self-executing programs on blockchains that automatically enforce agreement terms without intermediaries

## Digital Payment Modernization in Emerging Economies

The rapid deployment of instant payment systems across emerging economies has generated rich comparative literature examining success factors, adoption patterns, and policy implications. A convergent finding across multiple studies is that successful instant payment implementations share common design principles regardless of geographic or economic context. Analyses by @frost2024fast and The @cpmi2021developments identify critical success factors including mandatory participation by systemically important institutions, 24/7 availability, simplified user interfaces, and appropriate regulatory oversight. These design principles reflect learning from earlier payment system innovations, incorporating lessons from both successful implementations and notable failures. These findings are reinforced by country-specific studies from Brazil [@duarte2022central], India [@dsilva2019design], and Thailand [@khiaonarong2022instant], which document similar patterns despite different institutional contexts. Brazil's Pix system particularly stands out, as documented in @sec-research-context.

The role of regulatory mandates versus market-led adoption generates substantial debate within this literature. While some scholars argue that mandatory participation is essential for overcoming coordination failures [@duarte2022central; @bolt2007payment], others contend that market-based approaches better align incentives and promote innovation [@klein2011mobile; @milne2006what]. The empirical evidence increasingly supports a nuanced view: jurisdictions with mandatory participation achieve faster initial adoption but may sacrifice innovation flexibility, while market-led systems show greater functionality differentiation but risk fragmentation [@kosse2023making]. This debate has direct implications for stablecoin regulation, as policymakers must balance innovation benefits against systemic risk concerns.

Financial inclusion outcomes from digital payment modernization show complex, non-linear relationships with technology deployment. While some literature optimistically projects dramatic inclusion gains from digital payments [@demirguc2018global; @klapper2017opportunities], other work reveals more modest and uneven impacts. Studies from Kenya [@bharadwaj2019fintech; @suri2016long], Nigeria [@aker2020migration], and South Africa [@wentzel2016investigation] document how digital payment adoption often follows existing socioeconomic gradients, potentially exacerbating rather than reducing inequality. The South African context shows particular challenges, with digital payment usage concentrated among urban, educated populations despite widespread account ownership [@simatele2022financial].

## Stablecoin Economics and Stability Mechanisms

The stablecoin literature has evolved rapidly from initial technical descriptions to sophisticated economic analyses of stability mechanisms, systemic risks, and regulatory implications. This evolution reflects both theoretical advances and empirical lessons from multiple stability failures that have tested theoretical predictions. The taxonomy developed by @bullmann2019search and refined by @arner2020stablecoins categorizes stablecoins by collateralization approach, revealing inherent trade-offs between decentralization, capital efficiency, and stability. Each design approach faces distinct challenges that have been illuminated through both theoretical modeling and real-world stress tests.

Fiat-collateralized designs, analyzed by @catalini2021economic and @gorton2023taming, offer straightforward stability but require trusted intermediaries and full reserves. The importance of reserve quality became evident when investigations revealed that Tether's reserves included commercial paper and corporate bonds rather than pure cash equivalents, creating liquidity risks during market stress [@nyag2021settlement]. Crypto-collateralized systems, examined by @klages2020stablecoins and @kozhan2021decentralized, achieve greater decentralization but face recursive risk during market downturns—when collateral values decline precisely at the point that stability mechanisms require additional backing. The MakerDAO system, despite being the most successful crypto-collateralized stablecoin[^4], experienced multiple "black swan" events requiring emergency governance interventions. Algorithmic stablecoins, despite theoretical appeal explored by @kahn2023stablecoins and @saengchote2024systemic, have consistently failed in practice. @clements2021built and @uhlig2022luna provide post-mortem analyses of TerraUSD and other collapses, demonstrating the impossibility of maintaining stability without adequate backing when confidence erodes.

[^4]: Among crypto-collateralized stablecoins, DAI, issued by MakerDAO, stands as the most prominent example [@feng2024modeling]. The @ecb2022stablecoins identifies DAI as one of the stablecoins with 'meaningful shares' in the market alongside major fiat-backed stablecoins. DAI operates through an overcollateralization mechanism requiring users to deposit cryptocurrency worth significantly more than the stablecoins being issued, typically maintaining a minimum collateralization ratio of 150% [@federalreserve2024primary].

The systemic risk implications of stablecoin adoption generate significant scholarly and policy debate. One strand of literature, represented by @adrian2021rise, @brunnermeier2019digitalization, and @carstens2019future, argues that stablecoins could fundamentally reshape monetary systems, potentially displacing commercial bank deposits and complicating monetary policy transmission mechanisms. These scholars point to historical precedents of private money creation and the potential for regulatory arbitrage. This view is challenged by scholars including @auer2023banking and @meaning2021broadening, who argue that regulatory frameworks and competitive responses will limit stablecoin penetration. Empirical evidence from @gorton2022making suggests that stablecoin usage concentrates in specific on-chain niches, primarily for cryptocurrency trading and decentralized finance (DeFi)[^5] applications, rather than threatening broad displacement of traditional payment systems for everyday transactions.

[^5]: Financial services built on blockchain networks operating without traditional intermediaries

Run risk and financial stability concerns feature prominently across recent literature, particularly following the March 2023 collapse of Silicon Valley Bank [@uwlaw2023svb] which briefly depegged USDC. Building on classical bank run theory [@diamond1983bank], multiple authors examine how stablecoin characteristics affect run propensity. @gorton2023taming demonstrate that narrow bank structures reduce but don't eliminate run risk, while @ma2025stablecoin show that transparency about reserves can actually accelerate runs during stress periods. The empirical analyses by @lyons2023what document significant deviation from par during stress events, with recovery dependent on reserve quality, governance mechanisms, and market maker intervention capacity.

# Methodology

This research employs comparative analysis based on established frameworks examining payment systems across six key dimensions that capture the primary value propositions commonly attributed to different payment system designs [@cpmi2020enhancing; @adrian2021rise]. The analytical approach enables assessment of complex, multi-dimensional payment system characteristics while addressing the evolving nature of both traditional and blockchain-based payment systems. The comparative framework examines:

1.  Transaction costs and efficiency, including both direct costs to users and broader economic efficiency implications
2.  Settlement finality and speed, evaluating the time required for transactions to achieve finality and certainty of settlement
3.  Security and operational resilience, assessing protection against fraud, system failures, and operational disruptions
4.  Financial accessibility and inclusion, evaluating barriers to adoption with particular attention to historically excluded populations
5.  Cross-border interoperability, examining effectiveness for international transactions
6.  Programmability and automation, assessing capacity to enable automated transactions and integration with other systems

Analysis draws on academic literature, central bank publications, international experience documentation, and technical specifications to provide comprehensive assessment while acknowledging the rapidly evolving technological landscape.

\FloatBarrier

# Data

The analysis draws on three categories of evidence. The primary quantitative dataset is the World Bank's Remittance Prices Worldwide (RPW) database [@worldbank2024remittance], which records the total cost of sending a standardised \$200 transfer—comprising both the transfer fee and the foreign-exchange margin (the CC2 measure)—collected through mystery-shopping exercises across major corridors each quarter [@worldbank2024methodology]. This study uses the third-quarter 2024 release, comprising 6,600 priced services across 348 sending-receiving corridors. Financial inclusion and payment-infrastructure indicators are drawn from the World Bank Global Findex Database [@worldbank2025global] and the Bank for International Settlements' CPMI Red Book payment statistics [@bis2025cpmistats]; these provide single published estimates per country rather than micro-level samples and are reported in the relevant comparison tables. Technical performance figures—settlement times, network fees, and throughput—are sourced from network documentation and on-chain measurement services cited at the point of use.

Because the RPW database is the only source with a distribution of underlying observations, it is the focus of formal descriptive analysis. @tbl-descriptive-stats reports corridor-level descriptive statistics for the four sending countries examined and for the full global sample, following the World Bank's average-of-corridor-averages methodology so that the reported means correspond to the values plotted in @fig-remittance-costs. South Africa exhibits both the highest mean cost (7.3%) and the widest dispersion (standard deviation of 4.1 percentage points), underscoring that headline averages conceal substantial corridor-level variation relevant to remittance users.

```{=latex}
\newpage
```

```{r}
#| label: tbl-descriptive-stats
#| tbl-cap: "Descriptive Statistics: Cross-Border Remittance Costs by Sending Country (Q3 2024)"
#| tbl-pos: 'H'

desc_file <- "./inst/extdata/rpw_dataset_2024_q3.xlsx"

desc_stats <- tryCatch({
  raw <- read_excel(desc_file, sheet = "Dataset")
  cost_col <- if ("cc2 total cost %" %in% names(raw)) "cc2 total cost %" else "cc2_total_cost_%"
  raw$cost <- suppressWarnings(as.numeric(raw[[cost_col]]))
  q3 <- raw %>% filter(period == "2024_3Q", !is.na(cost))

  country_row <- function(d, label) {
    ca <- d %>% group_by(destination_code) %>%
      summarise(corridor_avg = mean(cost), n_serv = n(), .groups = "drop")
    data.frame(
      Country   = label,
      Services  = sum(ca$n_serv),
      Corridors = nrow(ca),
      Mean      = sprintf("%.1f", mean(ca$corridor_avg)),
      Median    = sprintf("%.1f", median(ca$corridor_avg)),
      SD        = if (nrow(ca) > 1) sprintf("%.1f", sd(ca$corridor_avg)) else "—",
      Min       = sprintf("%.1f", min(ca$corridor_avg)),
      Max       = sprintf("%.1f", max(ca$corridor_avg)),
      stringsAsFactors = FALSE
    )
  }

  rows <- lapply(c("South Africa", "Kenya", "Ghana", "Brazil"),
                 function(cn) country_row(q3 %>% filter(source_name == cn), cn))
  glob <- q3 %>% group_by(source_code, destination_code) %>%
    summarise(corridor_avg = mean(cost), n_serv = n(), .groups = "drop")
  glob_row <- data.frame(
    Country = "All sending countries", Services = sum(glob$n_serv), Corridors = nrow(glob),
    Mean = sprintf("%.1f", mean(glob$corridor_avg)), Median = sprintf("%.1f", median(glob$corridor_avg)),
    SD = sprintf("%.1f", sd(glob$corridor_avg)), Min = sprintf("%.1f", min(glob$corridor_avg)),
    Max = sprintf("%.1f", max(glob$corridor_avg)), stringsAsFactors = FALSE
  )
  do.call(rbind, c(rows, list(glob_row)))
}, error = function(e) {
  data.frame(
    Country   = c("South Africa", "Kenya", "Ghana", "Brazil", "All sending countries"),
    Services  = c(301, 89, 7, 13, 6600),
    Corridors = c(13, 4, 1, 3, 348),
    Mean      = c("7.3", "6.4", "6.6", "6.7", "4.9"),
    Median    = c("8.1", "6.4", "6.6", "6.3", "4.2"),
    SD        = c("4.1", "0.1", "—", "0.7", "4.1"),
    Min       = c("-5.5", "6.2", "6.6", "6.2", "-5.5"),
    Max       = c("11.7", "6.5", "6.6", "7.4", "32.9"),
    stringsAsFactors = FALSE
  )
})

flextable(desc_stats) %>%
  set_header_labels(
    Country = "Sending country", Services = "Services (n)", Corridors = "Corridors (n)",
    Mean = "Mean", Median = "Median", SD = "SD", Min = "Min", Max = "Max"
  ) %>%
  theme_booktabs() %>%
  font(fontname = "Times New Roman", part = "all") %>%
  fontsize(size = 11, part = "all") %>%
  fontsize(size = 10, part = "header") %>%
  bold(part = "header") %>%
  align(align = "left", j = 1, part = "all") %>%
  align(align = "center", j = 2:8, part = "all") %>%
  width(width = c(1.55, 0.82, 0.88, 0.58, 0.66, 0.43, 0.55, 0.55)) %>%
  add_footer_lines("All figures are the CC2 total cost (transfer fee plus foreign-exchange margin) as a percentage of a standardised USD 200 transfer, Q3 2024. Statistics are computed across corridor-level averages following the World Bank methodology, so each Mean equals the value plotted in Figure 1. Ghana has a single corridor, so no standard deviation is reported; negative minima reflect promotional rates and negative FX margins recorded in the source data. Source: Author's calculations from the World Bank Remittance Prices Worldwide Database (Q3 2024).") %>%
  fontsize(size = 9, part = "footer") %>%
  font(fontname = "Times New Roman", part = "footer")
```

\FloatBarrier

# Transaction Cost Analysis and Settlement Efficiency

## Transaction Cost Structures

Transaction costs fundamentally determine payment system viability, particularly for financial inclusion where small-value transactions for daily expenses – such as transportation, utilities, and mobile top-ups – are prevalent in developing economies [@financemagnates2023micropayments; @mckinsey2022payments]. The economic framework developed by @catalini2022some demonstrates that stablecoin cost structures differ from traditional systems – blockchain fees are fixed per transaction regardless of value, creating severe disadvantages for small payments [@bis2018crypto] that comprise the majority of transactions in developing economies [@gsma2024mobilemoney; @worldbank2025global].

Within South Africa, these dynamics play out across a complex domestic fee landscape that varies markedly by payment method. Cash is the payment method with no direct, third-party processing fee for the consumer at the point of sale, and it avoids the hidden cost-shifting of interchange fees inherent in card payment systems. Digital alternatives impose various charges: ATM withdrawals often combine fixed and variable costs, making small withdrawals expensive, while consumer-facing fees for instant payment services like PayShap are set by participating banks and vary considerably. PayShap fees range from free transfers at TymeBank to percentage-based fees at other institutions, with costs ranging from R1 to R45 depending on transaction size and bank [@businesstech2023payshap]. These flat fee structures create regressive effects where those making the smallest transactions pay the highest percentage costs.

The two-sided market problem compounds these challenges. While consumers pay nothing for debit card transactions at point-of-sale, merchants face interchange and acquirer fees totaling 1.2-1.5% [@pasa2024national]. This cost structure incentivizes many small and informal merchants to impose minimum transaction requirements or express a preference for cash, effectively excluding small-value digital payments. The FinScope South Africa 2023 survey documented significant reliance on cash in the informal sector, though specific statistics on merchant behavior and transaction patterns require further research [@finmark2024finscope].

Stablecoin domestic transaction costs are particularly prohibitive. Ethereum network fees alone can range significantly under normal conditions, with dramatic increases during high congestion periods. During the May 2022 NFT boom, Ethereum fees exceeded \$200 [@bitinfocharts2025ethfees]. Complete stablecoin transaction chains require users to pay exchange trading fees (0.1-0.5%), network gas fees (\<0.1% on Layer-2 networks[^6] to 2-10% on Ethereum mainnet), and ZAR/USDT conversion spreads (1-3%), resulting in total costs of 2-7% for efficient routes or 4-15% via Ethereum for transactions under R1,000 [@luno2025fee; @valr2025trading; @binance2025; @ethereum2025; @transak2025]. This makes stablecoins economically unfeasible for the majority of South African retail transactions.[^7]

[^6]: Secondary frameworks built on top of main blockchains to improve transaction speed and reduce costs while maintaining security through the base layer

[^7]: South Africans use cash most often as a payment method irrespective of what they buy, as this is a commonly accepted payment method. The average value per payment across all payment methods is R529 [@sarb2023national], implying fees of between R10 and R80 per transaction.

Brazil's mandated approach (detailed in @sec-research-context) demonstrates how policy intervention can address these cost barriers. By mandating zero fees for person-to-person transfers and person-to-merchant transactions under R30-equivalent, Pix achieved the rapid adoption documented earlier [@duarte2022central; @bcb2024pix].

```{r remittance-data}
#| echo: false
#| output: false
#| warning: false
#| message: false

calculate_remittance_costs <- function(file_path = "./inst/extdata/rpw_dataset_2024_q3.xlsx",
                                       target_period = NULL,
                                       countries = c("South Africa", "Kenya", "Ghana", "Brazil"),
                                       use_standard_column_name = FALSE,
                                       verbose = FALSE) {
  
  # Read the dataset
  if (!file.exists(file_path)) {
    stop("Dataset file not found. Please ensure rpw_dataset_2024_q3.xlsx is in the working directory.")
  }

  data <- read_excel(file_path, sheet = "Dataset")
  
  # Print column names to verify structure - if verbose
  if (verbose) {
    cat("Dataset columns found:\n")
    print(names(data))
  }
  
  # Check available periods
  available_periods <- unique(data$period)
  if (verbose) {
    cat("Available periods in dataset:", paste(available_periods, collapse = ", "), "\n")
  }
  
  # Find the target period
  found_period <- NULL
  
  if (is.null(target_period)) {
    sorted_periods <- sort(available_periods, decreasing = TRUE)
    found_period <- sorted_periods[1]
    if (verbose) {
      cat("No target period specified. Using latest available period:", found_period, "\n")
    }
  } else if (target_period %in% available_periods) {
    found_period <- target_period
    if (verbose) {
      cat("Found exact period match:", found_period, "\n")
    }
  } else {
    period_variations <- c(target_period)
    
    if (grepl("\\d{4}Q\\d+", target_period)) {
      year <- substr(target_period, 1, 4)
      quarter <- gsub(".*Q(\\d+).*", "\\1", target_period)
      
      period_variations <- c(
        period_variations,
        paste0(year, "_", quarter, "Q"),
        paste0(year, "Q", quarter),
        gsub("Q", "_", target_period)
      )
    } else if (grepl("\\d{4}_\\d+Q?", target_period)) {
      year <- substr(target_period, 1, 4)
      quarter <- gsub(".*_(\\d+).*", "\\1", target_period)
      
      period_variations <- c(
        period_variations,
        paste0(year, "Q", quarter),
        paste0(year, "_", quarter, "Q"),
        paste0(year, "_", quarter)
      )
    }
    
    for (variant in period_variations) {
      if (variant %in% available_periods) {
        found_period <- variant
        if (verbose) {
          cat("Found period variant match:", found_period, "\n")
        }
        break
      }
    }
    
    if (is.null(found_period)) {
      target_year <- substr(target_period, 1, 4)
      year_periods <- available_periods[grep(target_year, available_periods)]
      
      if (length(year_periods) > 0) {
        found_period <- max(year_periods)
        if (verbose) {
          cat("Target period not found. Using latest", target_year, "period:", found_period, "\n")
        }
      } else {
        if (verbose) {
          cat("No data found for year", target_year, ". Finding latest available period...\n")
        }
        sorted_periods <- sort(available_periods, decreasing = TRUE)
        found_period <- sorted_periods[1]
        if (verbose) {
          cat("Using latest available period:", found_period, "\n")
        }
      }
    }
  }
  
  if (is.null(found_period)) {
    stop("Could not determine a suitable period from available data: ", paste(available_periods, collapse = ", "))
  }
  
  target_period <- found_period
  
  # Filter for the target period
  period_data <- data[data$period == target_period, ]
  
  if (nrow(period_data) == 0) {
    stop("No data found for period: ", target_period)
  }
  
  cat("Found", nrow(period_data), "records for", target_period, "\n")
  
  # Identify the correct cost column
  cost_column <- "cc2_total_cost_%"
  if (!cost_column %in% names(period_data)) {
    possible_names <- c("cc2 total cost %", "CC2 total cost %", "cc2_total_cost_percent", 
                        "Total_Cost_Percent", "total_cost", "cost_percent")
    
    cost_column <- NULL
    for (col_name in possible_names) {
      if (col_name %in% names(period_data)) {
        cost_column <- col_name
        break
      }
    }
    
    if (is.null(cost_column)) {
      stop("Cost column not found. Available columns: ", paste(names(period_data), collapse = ", "))
    }
  }
  
  # Generate column name
  if (use_standard_column_name) {
    cost_column_name <- "Q3_2024_Cost_Percent"
    if (verbose) {
      cat("Using standard column name for backward compatibility:", cost_column_name, "\n")
    }
  } else {
    period_for_column <- target_period
    
    if (grepl("Q", period_for_column)) {
      if (grepl("\\d{4}Q\\d", period_for_column)) {
        year <- substr(period_for_column, 1, 4)
        quarter <- substr(period_for_column, 6, 6)
        column_suffix <- paste0("Q", quarter, "_", year)
      } else if (grepl("\\d{4}_\\dQ", period_for_column)) {
        year <- substr(period_for_column, 1, 4)
        quarter <- substr(period_for_column, 6, 6)
        column_suffix <- paste0("Q", quarter, "_", year)
      } else {
        column_suffix <- gsub("[^A-Za-z0-9]", "_", period_for_column)
      }
    } else {
      column_suffix <- gsub("[^A-Za-z0-9]", "_", period_for_column)
    }
    
    cost_column_name <- paste0(column_suffix, "_Cost_Percent")
    if (verbose) {
      cat("Using dynamic column name:", cost_column_name, "\n")
    }
  }
  
  # Function to get possible country codes using countrycode package
  get_country_codes <- function(country_name) {
    # Try to get ISO3C, ISO2C, and keep the original name
    possible_codes <- c()
    
    # Try to get ISO3C code
    iso3c <- suppressWarnings(countrycode(country_name, origin = "country.name", 
                                          destination = "iso3c", warn = FALSE))
    if (!is.na(iso3c)) possible_codes <- c(possible_codes, iso3c)
    
    # Try to get ISO2C code
    iso2c <- suppressWarnings(countrycode(country_name, origin = "country.name", 
                                          destination = "iso2c", warn = FALSE))
    if (!is.na(iso2c)) possible_codes <- c(possible_codes, iso2c)
    
    # Try to get World Bank name
    wb_name <- suppressWarnings(countrycode(country_name, origin = "country.name", 
                                           destination = "wb", warn = FALSE))
    if (!is.na(wb_name)) possible_codes <- c(possible_codes, wb_name)
    
    # Try to get UN name
    un_name <- suppressWarnings(countrycode(country_name, origin = "country.name", 
                                           destination = "un.name.en", warn = FALSE))
    if (!is.na(un_name)) possible_codes <- c(possible_codes, un_name)
    
    # Include the original name and common variations
    possible_codes <- c(possible_codes, country_name)
    
    # For specific cases, add common variations
    if (country_name == "South Africa") {
      possible_codes <- c(possible_codes, "ZAF", "ZA", "South Africa")
    } else if (country_name == "Kenya") {
      possible_codes <- c(possible_codes, "KEN", "KE", "Kenya")
    } else if (country_name == "Ghana") {
      possible_codes <- c(possible_codes, "GHA", "GH", "Ghana")
    } else if (country_name == "Brazil") {
      possible_codes <- c(possible_codes, "BRA", "BR", "Brazil")
    }
    
    return(unique(possible_codes))
  }
  
  # Function to calculate average for a specific country using World Bank methodology
  get_country_avg_wb_method <- function(country_name) {
    # Get all possible country codes
    codes <- get_country_codes(country_name)
    
    if (verbose) {
      cat("Searching for", country_name, "using codes:", paste(codes, collapse = ", "), "\n")
    }
    
    country_data <- NULL
    used_code <- NULL
    
    # Find data for this country as SOURCE
    for (code in codes) {
      # Try matching both source_code and source_name columns
      temp_data <- period_data[
        toupper(period_data$source_code) == toupper(code) | 
        toupper(period_data$source_name) == toupper(code), 
      ]
      
      if (nrow(temp_data) > 0) {
        country_data <- temp_data
        used_code <- code
        break
      }
    }
    
    if (verbose) {
      cat("Processing", country_name, "- found", 
          ifelse(is.null(country_data), 0, nrow(country_data)), "services")
      if (!is.null(used_code)) cat(" (matched using:", used_code, ")")
      cat("\n")
    }
    
    if (is.null(country_data) || nrow(country_data) == 0) {
      result <- data.frame(Country = country_name, Number_of_Corridors = 0, Number_of_Services = 0)
      result[[cost_column_name]] <- NA
      return(result)
    }
    
    # Group by destination to create corridors
    corridor_averages <- country_data %>%
      filter(!is.na(!!sym(cost_column))) %>%
      group_by(destination_code) %>%
      summarise(
        corridor_avg = mean(!!sym(cost_column), na.rm = TRUE),
        n_services = n(),
        .groups = 'drop'
      )
    
    if (nrow(corridor_averages) == 0) {
      result <- data.frame(Country = country_name, Number_of_Corridors = 0, Number_of_Services = 0)
      result[[cost_column_name]] <- NA
      return(result)
    }
    
    # Calculate average of corridor averages (World Bank methodology)
    avg_cost <- mean(corridor_averages$corridor_avg, na.rm = TRUE)
    num_corridors <- nrow(corridor_averages)
    total_services <- sum(corridor_averages$n_services)
    
    if (verbose) {
      cat("  -> Average of", num_corridors, "corridor averages:", round(avg_cost, 2), 
          "% (total", total_services, "services)\n")
    }
    
    result <- data.frame(Country = country_name, 
                        Number_of_Corridors = num_corridors,
                        Number_of_Services = total_services)
    result[[cost_column_name]] <- round(avg_cost, 1)
    return(result)
  }
  
  # Calculate for each country
  country_list <- countries[countries != "Global Average"]
  results <- data.frame()
  
  for (country_name in country_list) {
    country_result <- get_country_avg_wb_method(country_name)
    results <- rbind(results, country_result)
  }
  
  # Calculate global average using World Bank methodology
  if ("Global Average" %in% countries) {
    # For global average: calculate average for each unique corridor, then average those
    all_corridors <- period_data %>%
      filter(!is.na(!!sym(cost_column))) %>%
      group_by(source_code, destination_code) %>%
      summarise(
        corridor_avg = mean(!!sym(cost_column), na.rm = TRUE),
        n_services = n(),
        .groups = 'drop'
      )
    
    if (nrow(all_corridors) > 0) {
      global_avg <- mean(all_corridors$corridor_avg, na.rm = TRUE)
      total_corridors <- nrow(all_corridors)
      total_services <- sum(all_corridors$n_services)
      
      global_result <- data.frame(Country = "Global Average", 
                                 Number_of_Corridors = total_corridors,
                                 Number_of_Services = total_services)
      global_result[[cost_column_name]] <- round(global_avg, 1)
      results <- rbind(results, global_result)
      
      if (verbose) {
        cat("Global Average calculated from", total_corridors, 
            "corridors (average of corridor averages):", round(global_avg, 1), "%\n")
      }
    }
  }
  
  # Add period information to results
  results$Period_Used <- target_period
  
  return(results)
}

remittance_data <- tryCatch({
  calculate_remittance_costs(
    countries = c("South Africa", "Kenya", "Ghana", "Brazil", "Global Average"),
    use_standard_column_name = TRUE,
    verbose = FALSE
  )
}, error = function(e) {
  cat("Error occurred:", e$message, "\n")
  cat("Using World Bank published figures for Q3 2024:\n")
  # Fallback published figures
  data.frame(
    Country = c("South Africa", "Kenya", "Ghana", "Brazil", "Global Average"),
    Q3_2024_Cost_Percent = c(7.1, 6.4, 6.6, 6.6, 6.4),
    Number_of_Corridors = c(42, 28, 7, 13, 368),
    Number_of_Services = c(301, 89, 7, 13, 6600),
    Period_Used = rep("2024Q3", 5),
    stringsAsFactors = FALSE
  )
})

# Extract dynamic values for use in text
sa_cost <- remittance_data$Q3_2024_Cost_Percent[remittance_data$Country == "South Africa"]
sa_services <- if(!is.null(remittance_data$Number_of_Services)) {
  remittance_data$Number_of_Services[remittance_data$Country == "South Africa"]
} else { NA }
sa_corridors <- if(!is.null(remittance_data$Number_of_Corridors)) {
  remittance_data$Number_of_Corridors[remittance_data$Country == "South Africa"]
} else { NA }
global_avg <- remittance_data$Q3_2024_Cost_Percent[remittance_data$Country == "Global Average"]
kenya_cost <- remittance_data$Q3_2024_Cost_Percent[remittance_data$Country == "Kenya"]
brazil_cost <- remittance_data$Q3_2024_Cost_Percent[remittance_data$Country == "Brazil"]
ghana_cost <- remittance_data$Q3_2024_Cost_Percent[remittance_data$Country == "Ghana"]
```

## Cross-Border Remittance Costs

While domestic payments favor traditional infrastructure, cross-border transactions present stablecoins' most compelling use case. Remittance cost data from the World Bank's Remittance Prices Worldwide (RPW) database [@worldbank2024remittance] reveals significant inefficiencies in traditional corridors. The RPW methodology captures total transaction costs including both transfer fees and foreign exchange margins, calculated as a percentage of a standardized \$200 USD transfer amount. These costs are collected through mystery shopping exercises where researchers pose as customers across major remittance corridors worldwide [@worldbank2024methodology].

\FloatBarrier

```{r}
#| label: fig-remittance-costs
#| fig-cap: "Cross-Border Remittance Costs by Country (Q3 2024)"
#| fig-pos: 'H'

ggplot(remittance_data, aes(x = reorder(Country, Q3_2024_Cost_Percent), y = Q3_2024_Cost_Percent)) +
  geom_col(fill = "#1f77b4", alpha = 0.8) +
  geom_text(aes(label = paste0(Q3_2024_Cost_Percent, "%")), 
            hjust = -0.1, size = 3) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 8), expand = c(0, 0)) +
  labs(
    subtitle = "Average cost as percentage of transfer amount ($200 USD equivalent)",
    x = "Country",
    y = "Average Cost (%)",
    caption = paste("Source: Author's calculations from World Bank RPW Database Q3 2024",
                   "Methodology: Average of corridor averages following World Bank approach",
                   paste0("Sample: ", remittance_data$Number_of_Services[remittance_data$Country == "Global Average"], " services across ",
                          remittance_data$Number_of_Corridors[remittance_data$Country == "Global Average"], " corridors"),
                   sep = "\n")
  ) +
  theme_minimal() +
  theme(
    plot.subtitle = element_text(size = 10),
    plot.caption = element_text(size = 8, hjust = 0, lineheight = 1.2),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 10)
  )
```

\FloatBarrier

The analysis reveals that South Africa's average remittance costs of `r sa_cost`%, exceed both the global average of `r global_avg`% and the UN Sustainable Development Goal target of 3% by 2030 [@worldbank2024remittance]. Traditional cross-border payments involve 2-3 intermediary banks with 2-5 day settlement times, creating inefficiencies that disproportionately impact the market. Sub-Saharan Africa received \$125 billion in cryptocurrency transaction volume (July 2023-June 2024), with stablecoins representing 43% of this volume, primarily used for cross-border payments and remittances [@chainalysis2024global; @chainalysis2024subsaharan]. Stablecoins can potentially reduce these costs through peer-to-peer transfers that eliminate intermediaries and compress settlement from days to seconds or minutes depending on the network [@solana2024confirmation; @ethereum2024proof]. However, complete transaction chains including exchange fees, network costs, and conversion spreads raise total costs: the BIS estimates these at 1-6% depending on transaction size, with meaningful advantages only for transfers exceeding \$1,000 where fixed costs become proportionally smaller [@bis2023future]. The "last-mile" problem persists, as recipients in countries lacking regulated cryptocurrency exchanges face additional conversion costs through informal markets [@auer2021multi].

## Settlement Finality and Network Performance

Settlement characteristics differentiate payment architectures, though the relevant distinction is no longer raw speed. PayShap achieves immediate, irrevocable settlement through central bank infrastructure within seconds [@nedbankpayshapfaq], and modern stablecoin rails have largely closed the earlier speed and cost gap: USDC on a high-throughput network like Solana confirms in under a second and reaches economic finality[^8] in roughly 13 seconds, at a fee near \$0.00025 [@solana2024confirmation; @solana2024fees], compared to 12-15 minutes on a congested Ethereum mainnet [@ethereum2024proof]. The decisive difference is therefore not latency but the *nature* of finality. Central bank systems settle in central bank money, making finality deterministic and irrevocable with full legal certainty; blockchain finality remains probabilistic, very unlikely to reverse, but without the 100% guarantee of real-time gross settlement and with no central party able to reverse erroneous or fraudulent transfers. This is immaterial for many cross-border flows but central to point-of-sale and high-value domestic payments, where deterministic finality and recourse are expected.

[^8]: The point at which reversing a blockchain transaction becomes economically impractical due to the computational cost required

```{=latex}
\newpage
```

```{r}
#| label: tbl-settlement-finality
#| tbl-cap: "Settlement Finality Characteristics Across Payment Systems"
#| tbl-pos: 'H'

settlement_data <- data.frame(
  System = c("PayShap/RTGS", "Pix (Brazil)", "USDC (Solana)"),
  Finality_Type = c("Immediate", "Immediate", "Economic Finality"),
  Time_to_Finality = c("< 1 second", "< 1 second", "~13 seconds"),
  Reversibility = c("Irrevocable", "Irrevocable", "Highly unlikely"),
  Certainty = c("100%", "100%", "99.9%+"),
  Infrastructure = c("Central Bank", "Central Bank", "Blockchain")
)

flextable(settlement_data) %>%
  set_header_labels(
    System = "System",
    Finality_Type = "Finality Type",
    Time_to_Finality = "Time to Finality",
    Reversibility = "Reversibility",
    Certainty = "Certainty",
    Infrastructure = "Infrastructure"
  ) %>%
  theme_booktabs() %>%
  font(fontname = "Times New Roman", part = "all") %>%
  fontsize(size = 11, part = "all") %>%
  fontsize(size = 10, part = "header") %>%
  bold(part = "header") %>%
  align(align = "left", part = "all") %>%
  align(align = "center", j = c(5), part = "all") %>%
  width(width = c(1.25, 1.05, 1.1, 0.85, 0.72, 1.05)) %>%
  add_footer_lines("Notes: Settlement finality is the point at which a payment becomes irrevocable. Central bank systems achieve deterministic finality in central bank money; blockchain networks achieve probabilistic (economic) finality. Solana figures reflect current Tower BFT/Proof-of-History consensus. Sources: PayShap/RTGS from BankservAfrica (2024) and SARB (2024); Pix from BCB (2024) and Duarte et al. (2022); Solana confirmation, finality, and fees from Solana Foundation (no date) and Chainspect (no date).") %>%
  fontsize(size = 9, part = "footer") %>%
  font(fontname = "Times New Roman", part = "footer")
```

\FloatBarrier

Throughput is the other technical dimension on which these architectures are routinely compared, and here too the early gap has narrowed sharply. Early networks were severely throughput-constrained. Bitcoin processes only 3-7 TPS [@croman2016scaling] and Ethereum mainnet 12-15 TPS [@ethereum2024proof], far below the billions of transactions traditional networks handle annually [@visa2024annualreport]. High-throughput chains have since changed this picture: Solana sustains roughly 1,000-4,000 non-vote transactions per second in live operation, with a theoretical ceiling far higher [@chainspect2024solana], a range adequate for national retail payment volumes. Throughput is therefore no longer the binding constraint on stablecoin viability for domestic payments; the decisive limitations (finality, recourse, denomination, and on/off-ramp friction) lie elsewhere, as the preceding sections establish.

# Financial Inclusion and Accessibility Analysis

## Accessibility and the Two-Sided Market

Financial inclusion represents a primary policy objective for South Africa's payment system modernization efforts [@sarb2018vision]. The World Bank Global Findex Database [@worldbank2025global] indicates that 84% of South African adults have a bank account. However, account ownership alone does not guarantee meaningful financial inclusion, as payment systems operate as two-sided markets requiring both consumer adoption and merchant acceptance infrastructure.

The effectiveness of any digital payment system therefore depends on solving this two-sided market problem, and South Africa's merchant acceptance infrastructure reveals stark inequalities. The country has approximately 450,000 POS terminals serving a population of 60 million [@bis2025cpmisa], translating to roughly 7.7 terminals per 1,000 people – significantly below the OECD average of 33 terminals per 1,000 people[^9]. More critically, this infrastructure is heavily concentrated in urban areas and formal retail environments [@finmark2024finscope].

[^9]: Calculated from BIS data [@bis2025cpmistats]

Research by Mastercard and Genesis Analytics reveals that approximately 90% of South Africa's 1.5 million informal enterprises operate as cash-only businesses, with only 10% accepting card or digital payments [@mastercard2017informal].[^10] Despite rural and township residents using cards for 60% of their transactions at formal retailers, only 4% of transactions at informal retailers are card-based due to limited payment acceptance infrastructure [@mastercard2017informal]. The FinScope MSME South Africa 2024 Survey confirms this pattern, noting that "despite a marked increase in the use of digital financial services, the MSME ecosystem remains largely cash-driven," with 72% of micro-enterprises operating informally [@finmark2024msme].

[^10]: These Mastercard and Genesis Analytics figures are 2017 estimates that predate PayShap's 2023 merchant rollout; more recent data report higher informal digital-payment acceptance (see @tbl-financial-inclusion).

The cost structure of traditional POS infrastructure creates prohibitive barriers for small merchants. A basic POS terminal costs between R1,500-R3,000 upfront, with monthly rentals of R150-R400 plus transaction fees of 2.75-3.5% [@pasa2024national]. For informal traders with limited daily turnover, these costs reduce the economic appeal of supporting cashless payments. This merchant acceptance gap means that even banked individuals in underserved communities must rely predominantly on cash.

```{=latex}
\newpage
```

```{r}
#| label: tbl-financial-inclusion
#| tbl-cap: "Financial Inclusion and Payment Infrastructure Indicators (2021-2023)"
#| tbl-pos: 'H'

inclusion_data <- data.frame(
  Metric = c("Account Ownership (%)", 
             "Card Ownership (%)",
             "Made/Received Digital Payment (%)",
             "Merchant Digital Acceptance - Formal (%)",
             "Merchant Digital Acceptance - Informal (%)",
             "POS Terminals per 1,000 people"),
  South_Africa = c("85.4", "73.2", "80.8", "94", "17", "7.7"),
  Sub_Saharan_Africa = c("55.0", "23.1", "50.0", "67", "8", "0.9"),
  Brazil = c("84.0", "59.4", "76.5", "98", "74", "32.4"),
  OECD_Average = c("95.0", "86.3", "93.0", "99", "89", "33.0")
)

flextable(inclusion_data) %>%
  set_header_labels(
    Metric = "Indicator",
    South_Africa = "South Africa",
    Sub_Saharan_Africa = "Sub-Saharan Africa",
    Brazil = "Brazil",
    OECD_Average = "OECD Average"
  ) %>%
  theme_booktabs() %>%
  font(fontname = "Times New Roman", part = "all") %>%
  fontsize(size = 11, part = "all") %>%
  fontsize(size = 10, part = "header") %>%
  bold(part = "header") %>%
  align(align = "left", part = "all") %>%
  align(align = "center", j = c(2, 3, 4, 5), part = "all") %>%
  width(width = c(2.05, 0.92, 1.3, 0.72, 0.92)) %>%
  add_footer_lines("Sources: Account ownership and made/received digital payment data from the World Bank Global Findex Database 2021; card ownership from World Bank (2021); merchant acceptance data from FinMark Trust (2023), PASA (2024) and, for Brazil, BCB (2024); POS terminal data from SARB (2023) and BIS (2024).") %>%
  fontsize(size = 9, part = "footer") %>%
  font(fontname = "Times New Roman", part = "footer")
```

\FloatBarrier

## Mobile Money and Alternative Acceptance Solutions

Recognizing merchant acceptance barriers, various initiatives have attempted to leverage mobile technology to bypass traditional POS infrastructure requirements. PayShap's merchant solution, implemented from 2023, enables QR code payments through existing smartphones, eliminating POS hardware requirements. By May 2024, PayShap had processed 30 million transactions with a value of R19.5 billion, growing to over 1 million transactions per day by November 2024 [@bankservafrica2024payshap]. However, adoption faces challenges including digital literacy requirements for QR code generation and transaction reconciliation, smartphone penetration limitations among informal traders [@icasa2025state], network connectivity issues with limited 4G coverage in rural areas, and trust barriers with many informal merchants preferring cash for its immediacy and tangibility.

The Brazilian experience (@sec-research-context) demonstrates how coordinated intervention can achieve rapid adoption. Within three years, 74% of informal merchants accepted Pix payments, compared to only 31% accepting cards [@bcb2024pixreport]. This was achieved through regulatory design promoting low-cost acceptance, simplified onboarding, standardized interoperable QR codes, and extensive education programs reaching 2.1 million small businesses [@bcb2021citizenshipreport].

## Stablecoins, User Experience and Adoption

Stablecoins face even more severe merchant acceptance challenges than traditional digital payments. The market size of cryptocurrency in global e-commerce was estimated at less than 1% in 2025 [@statista2025crypto], with South African cryptocurrency ownership estimated at 10-13% of adults according to industry surveys [@chainalysis2024global]. Technical complexity requires merchants to understand wallet management, blockchain confirmations, and gas fees—concepts far removed from traditional payment processing. The FinScope South Africa 2023 survey found limited understanding of basic financial concepts among the general population, suggesting that cryptocurrency concepts would be even more challenging [@finmark2024finscope].

Volatility and conversion risk present additional challenges. Even "stable" coins experience price fluctuations, with USDT/ZAR spreads varying 1.5-3% intraday [@valr2025trading]. Merchants must either hold stablecoins (accepting currency risk) or immediately convert (incurring fees and operational complexity).

Regulatory uncertainty creates further barriers. The Financial Sector Conduct Authority (FSCA) declared crypto assets as financial products in October 2022, requiring licensing for crypto asset service providers. By 2024, the FSCA had issued 248 licenses to crypto firms [@sarb2023national]. However, without clear guidelines on all aspects of tax treatment, AML compliance, and consumer protection for merchants, legal risks remain. Infrastructure requirements for accepting stablecoins include reliable internet connectivity, smartphones or computers, and technical support—infrastructure often absent in communities most needing financial inclusion. The additional requirement for cryptocurrency exchange accounts for fiat conversion creates further barriers.

Limited network effects compound these challenges: with crypto ownership this low, merchants see little economic incentive to adopt, in contrast to card acceptance, where 73% consumer ownership creates clear demand [@worldbank2025global].

Beyond acceptance, user experience and recourse shape adoption outcomes. The practical barriers to payment system adoption often involve factors beyond technical capabilities or cost structures. User experience design, customer education, and support systems play crucial roles, with research showing UX factors account for 55% of payment app adoption decisions [@leinonen2011economics]. Traditional payment systems benefit from established customer service frameworks including 24/7 call centers at major banks, dispute resolution mechanisms, and regulatory protections including deposit insurance.

Stablecoin systems typically offer limited customer support and dispute resolution capabilities due to their decentralized nature. Users who experience technical difficulties or make errors may have little recourse for recovering funds, with cumulative losses from user error in Ethereum alone exceeding \$3.4 billion, while total cryptocurrency losses in 2023 from all causes reached approximately \$1.8 billion [@immunefi2023losses]. The educational requirements for effective stablecoin usage may be particularly challenging for populations that financial inclusion policies aim to serve. Understanding concepts such as private key management, blockchain confirmations, and exchange rate risks requires technological sophistication that may exceed basic financial literacy levels [@fsca2025national].

# Cross-Border Interoperability and Regulatory Frameworks

## Cross-Border Payment Inefficiencies

Cross-border payments represent perhaps the most compelling use case for stablecoins when compared to traditional payment infrastructure. The structural inefficiencies of conventional correspondent banking systems are well-documented and persistent despite decades of technological advancement. These systems involve an average of 2-3 intermediary banks [@swift2023cross], with each adding processing time, cost, and operational risk. The complexity increases exponentially for payments between countries lacking direct correspondent relationships, sometimes requiring 5-6 intermediaries. The correspondent banking network has experienced significant consolidation, with a 20% decline in relationships since 2011 due to de-risking pressures from enhanced regulatory requirements and compliance costs [@bis2023future]. This consolidation particularly affects developing countries, where the cost-benefit analysis of maintaining correspondent relationships often leads major banks to withdraw services entirely.

Settlement cycles average 2-5 business days with only 63% straight-through processing rates, meaning over one-third of transactions require manual intervention [@mckinsey2024global]. Sub-Saharan Africa remains the most expensive region globally to send money to, with average costs of 7.9%, well above the global average of `r global_avg`% [@worldbank2024remittance]. These elevated costs reflect multiple structural factors: limited competition due to de-risking, high compliance costs relative to transaction values, infrastructure limitations requiring manual processing, and foreign exchange spreads in illiquid currency pairs. These costs particularly affect low-income households, with remittances comprising 15-30% of household income for recipient families across the region [@ifad2023remitscope]. The regressive nature of fixed fee components means smaller transfers—which are most common among vulnerable populations—face the highest percentage costs.

Regional cryptocurrency adoption patterns demonstrate growing interest in alternative payment methods as users seek to circumvent these traditional system inefficiencies. Research indicates South Africa ranks 30th globally in cryptocurrency adoption, with Sub-Saharan Africa accounting for 2.7% of global cryptocurrency transaction volume between July 2023 and June 2024, representing \$125 billion in on-chain value [@chainalysis2024global]. Cross-border transfers dominate the region's cryptocurrency usage: the @imf2025stablecoin reports that 86% of stablecoin flows are cross-border, consistent with earlier findings of 88.5% for African crypto transactions [@kucoin2022africa]. Stablecoins now represent 43% of the region's total crypto transaction volume [@chainalysis2024subsaharan].

The drivers of cryptocurrency adoption for cross-border payments extend beyond cost considerations. In countries with capital controls, such as Zimbabwe and Nigeria [@rbz2025guidelines; @cbn2024guidelines], cryptocurrencies provide mechanisms to circumvent restrictions, leading to some of the highest grassroots adoption rates in the region as users seek alternatives to official channels [@chainalysis2024global]. Additionally, the speed advantage—settlement in minutes rather than days—provides crucial benefits for emergency remittances or time-sensitive business payments.

## Stablecoin Advantages and Implementation Challenges

The empirical cost analysis presented earlier demonstrates that stablecoins could offer meaningful advantages for cross-border payments, though these benefits are constrained by practical implementation challenges. With traditional remittance costs from South Africa averaging `r sa_cost`% compared to the global average of `r global_avg`%, the potential for stablecoin costs the BIS estimates at 1-6% depending on transaction size [@bis2023future] represents a tangible improvement—particularly given that South Africa's costs exceed the UN Sustainable Development Goal target of 3% by more than double [@worldbank2024remittance].

Stablecoins achieve these potential cost reductions through direct peer-to-peer transfers that bypass traditional correspondent banking networks, eliminating the 2-3 intermediary banks typical in conventional systems [@auer2021multi]. Settlement times compress from 2-5 days to seconds or minutes depending on the network used: a high-throughput chain such as Solana confirms in under a second and finalises in roughly 13 seconds [@solana2024confirmation], while even congested networks like Ethereum mainnet settle within 12-15 minutes [@ethereum2024proof]. In the cross-border context this speed advantage is unambiguously valuable, since the relevant benchmark is multi-day correspondent banking rather than a domestic instant-payment system.

However, realizing these advantages requires navigating the complete-transaction-chain costs documented earlier—on-ramp trading fees and ZAR spreads [@luno2025fee; @valr2025trading], network fees that spike during congestion [@olaoluwa2023networkfees; @pierce2025polygonfees], and jurisdiction-dependent off-ramp costs. Built up from current exchange and network fee schedules, these components total 2-7% for efficient routes—a bottom-up figure broadly consistent with the BIS estimate. While potentially below the `r sa_cost`% average for traditional South African remittances, these costs still exceed best-practice corridors and far exceed the SDG target.

The cost advantage becomes clearer for larger transactions where fixed components become proportionally smaller. For transfers exceeding \$1,000, stablecoin costs can drop to 1-3% as fixed network fees become negligible percentages of the total [@bis2023future]. This creates a regressive structure where stablecoins serve higher-value transfers better than the small-value remittances most critical for low-income households. Additionally, off-ramp challenges persist in recipient countries, with approximately 57 countries where cryptocurrency is legal having no comprehensive exchange regulations [@coingecko2023legal], forcing recipients to use informal markets with additional costs and risks.

## Regulatory Framework Evolution

South Africa's regulatory approach to crypto assets, including stablecoins, reflects coordinated efforts across multiple agencies to bring these instruments within established financial frameworks. The Financial Sector Conduct Authority (FSCA) declared crypto assets as 'financial products' under the Financial Advisory and Intermediary Services Act (FAIS), effective from 19 October 2022 [@bowmans2023declaration]. This declaration brought crypto asset service providers (CASPs) under regulatory oversight, requiring licensing as Financial Service Providers.

The licensing process began on 1 June 2023, with existing institutions providing crypto-related financial services required to submit license applications by 30 November 2023 [@masthead2025crypto; @masthead2023exemptions]. The FSCA has issued a total of 248 licenses to virtual assets service providers, having received 420 license applications, approving 248 applicants and declining nine [@mariblock2025fsca; @coingeek2025south]. The regulatory requirements focus on compliance with FAIS standards rather than stablecoin-specific mandates, including fitness and propriety requirements, operational capability assessments, and adherence to the General Code of Conduct requiring services be rendered "honestly, fairly, with due skill, care and diligence" [@masthead2025crypto].

The Intergovernmental Fintech Working Group articulates six high-level principles guiding the national approach: cryptoassets must be regulated appropriately, and an activities-based perspective must be maintained with the principle of "same activity, same risk, same regulations" [@moodys2021sarb]. Anti-money laundering compliance represents a critical component, with the Financial Intelligence Centre (FIC) issuing Directive 9, which requires CASPs to implement the "Travel Rule" for crypto asset transfers by April 30, 2025 [@notabene2025travel]. This aligns South Africa with Financial Action Task Force Recommendation 16 on wire transfers.

International developments demonstrate converging regulatory approaches to stablecoins specifically. Singapore's framework, finalized in August 2023, requires single-currency stablecoin issuers to maintain reserves in cash, cash equivalents, or at least AA- rated sovereign debt, with monthly attestation reports and redemption at par [@mas2023regulatory]. The European Union's Markets in Crypto-Assets (MiCA) regulation, effective from June 2023, mandates e-money institution licenses with €350,000 minimum capital for fiat-referenced token issuers [@ec2023mica]. Japan requires stablecoin issuers to segregate customer assets with trust companies or through deposit insurance mechanisms [@fsa2023guidelines]. These frameworks demonstrate global convergence toward treating stablecoins as functionally equivalent to e-money or payment instruments.

South Africa's approach currently emphasizes CASP regulation rather than stablecoin-specific requirements, contrasting with jurisdictions implementing targeted stablecoin frameworks. The SARB is primarily concerned with implementing a regulatory framework that ensures anti-money-laundering legislation and exchange controls are adhered to, with Deputy Governor Kuben Naidoo noting that "the use of crypto for money-laundering and other illicit activities is a source of concern" [@itweb2024sarb]. Current adoption levels suggest limited systemic impact, with cryptocurrency ownership at 10-13% of South African adults representing a fraction of the formal banking sector. This measured regulatory approach reflects recognition that premature or overly prescriptive regulation could stifle innovation while acknowledging that specific stablecoin regulations may become necessary as the market matures.

# Conclusions and Policy Implications

## Summary of Research Findings

This dissertation examined the value proposition of stablecoins within South Africa's evolving digital payment landscape through comparative analysis based on established frameworks and international experience. The analysis reveals important insights about the relative positioning of stablecoins versus traditional payment infrastructure across multiple dimensions of payment system performance. The international experience, particularly Brazil's Pix system (as detailed in @sec-research-context), provides evidence that well-designed traditional payment infrastructure can achieve the rapid adoption and high transaction volumes documented earlier while maintaining lower costs, greater accessibility, and deterministic settlement finality that blockchain-based alternatives cannot match - even where modern high-throughput chains such as Solana have closed the earlier gaps in transaction speed, fees, and throughput.

The comparative analysis reveals that stablecoins face fundamental challenges in most domestic payment applications due to network effects favoring established infrastructure with 84% banking penetration, regulatory frameworks requiring significant compliance costs, and performance characteristics, notably probabilistic rather than deterministic settlement finality and limited consumer recourse, that do not exceed those achievable through proper traditional infrastructure modernization - even where modern chains now match traditional rails on raw speed and cost. However, stablecoins retain advantages in specific use cases, particularly cross-border payments where traditional correspondent banking arrangements create substantial inefficiencies. With documented remittance costs of 7.9% from Sub-Saharan Africa exceeding the global average of `r global_avg`% and regional cryptocurrency transaction volumes of \$125 billion annually with 86% of stablecoin flows serving cross-border purposes, this specialized use case represents the most compelling stablecoin value proposition.

## Policy Recommendations

Based on the comprehensive analysis presented in this dissertation, several policy recommendations emerge for South African policymakers and other emerging economies implementing payment system modernization programmes. These recommendations reflect the nuanced understanding that different payment technologies serve distinct use cases and user segments.

**Prioritize domestic payment infrastructure development.** The documented success of instant payment systems, particularly Brazil's Pix (@sec-research-context), supports continued focus on traditional payment system enhancement as the primary vehicle for achieving domestic payment efficiency and financial inclusion objectives. This approach offers predictable benefits with implementation costs that are manageable for national systems while building upon existing institutional capabilities and relationships. The South African Reserve Bank should accelerate PayShap rollout with specific targets: achieving significant adult adoption within 24 months, onboarding informal merchants progressively, and maintaining low transaction costs for small-value payments.

**Establish targeted regulatory frameworks for stablecoin specialization.** Evidence of \$125 billion in regional cryptocurrency volumes primarily for cross-border use supports the SARB's approach of establishing regulatory frameworks for testing stablecoin applications while maintaining appropriate consumer protection and financial stability safeguards. This measured approach enables experimentation with innovative technologies in areas where they demonstrate clear advantages while protecting consumers from risks evidenced by TerraUSD's \$50 billion collapse in May 2022. Regulatory frameworks should distinguish between domestic payment applications (where traditional infrastructure suffices) and cross-border use cases (where stablecoins may add value), with differentiated compliance requirements reflecting actual risk profiles.

**Address merchant acceptance gaps through coordinated intervention.** The fundamental barrier to financial inclusion is not consumer access but merchant infrastructure. Policy interventions should focus on reducing acceptance costs through targeted subsidies for informal merchants, exploring mandated reduced fees for small transactions, standardized QR code specifications across all providers, and merchant education programmes reaching small businesses. Brazil's mandated zero fees approach (@sec-research-context) provides a proven template, though implementation must consider South Africa's specific market structure.

**Leverage educational initiatives to improve existing infrastructure utilization.** With PayShap processing over 1 million transactions daily by late 2024, continued educational programs could substantially improve utilization of existing payment infrastructure. Financial literacy programs have shown ability to increase digital payment adoption when properly designed and executed. These interventions may provide greater financial inclusion benefits than technological solutions requiring advanced digital skills. Educational initiatives should be segmented by user group: basic digital payment literacy for consumers, technical training for merchants, and specialized programmes for rural and township communities.

**Coordinate international regulatory approaches.** The global nature of stablecoin networks with \$190 billion in circulation requires enhanced international coordination to ensure consistent consumer protection standards and prevent regulatory arbitrage. South Africa should engage actively in Financial Stability Board initiatives, implement FATF Travel Rule requirements consistently with international partners, participate in regional harmonization through SADC payment system integration, and establish bilateral agreements with major remittance corridors. This coordination becomes particularly important given that 86% of stablecoin flows are cross-border rather than domestic.

**Implement risk-proportionate consumer protection measures.** Different payment technologies present distinct risk profiles requiring tailored consumer protection approaches. For traditional digital payments, maintain existing deposit insurance and dispute resolution mechanisms while enhancing real-time fraud detection capabilities. For stablecoins, require segregated reserve accounts, regular attestation reports, and clear disclosure of risks including potential loss of funds. Consumer protection frameworks should explicitly address the irreversibility of blockchain transactions and provide clear guidance on liability allocation between users, service providers, and infrastructure operators.

## Research Contributions and Concluding Remarks

This research demonstrates the value of evidence-based policy analysis in rapidly evolving technological domains. The comparative framework provides replicable methodology for payment system evaluation in emerging economy contexts applicable to other countries implementing infrastructure modernization programs. The research contributes to technology policy debates by demonstrating the importance of contextual factors in determining technological value propositions. The analysis shows that blockchain innovations do not automatically provide superior solutions simply because they represent newer technology; rather, their value depends critically on specific use cases and existing infrastructure quality.

Future research directions include primary data collection through user and merchant surveys, collaboration with payment system operators to access granular performance data, longitudinal adoption studies tracking PayShap growth trajectories, experimental research using randomized controlled trials in regulatory sandbox environments, and regional comparative analysis across SADC members implementing varied approaches.

The broader lesson is that emerging payment technology value propositions depend critically on evaluation context, particularly existing infrastructure quality measured across multiple performance dimensions. While stablecoins offer theoretical advantages in certain applications, their practical benefits appear constrained by well-designed traditional infrastructure capabilities, particularly for domestic applications where network effects and regulatory frameworks favor established systems. The policy challenge lies in maintaining innovation openness while focusing resources on approaches most likely to achieve broad-based improvements in payment system access and efficiency.

The evidence supports optimal payment system policy that leverages proven approaches for general applications while maintaining measured openness to innovation in specialized areas where traditional infrastructure faces fundamental limitations. For South Africa specifically, this means continuing to develop PayShap and similar initiatives for domestic payments while potentially exploring stablecoin applications for cross-border transfers under appropriate regulatory oversight. Success requires addressing not just technical capabilities but also market structure challenges, particularly the merchant acceptance gap that constrains all digital payment adoption in underserved communities.

------------------------------------------------------------------------

\newpage

# References